---
title: "Results"
output:
  pdf_document:
  html_document: default
header-includes:
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The $R_t$ Estimator Web App

We developed a Shiny web-app that allows users to estimate $R_t$ for the SARS-CoV-2 pandemic in all countries, all states in the United States,, and all counties in the United States with 95% confidence intervals (or credible intervals) for the estimations. The user can estimate R using 4 different methods:

1. A simple ratio
2. The method described by Cori et al. (2013)
3. The method described by Wallinga & Teunis (2004)
4. The method described by Wallinga & Lipsitch (2007)

How these 4 different methods are implemented are described in the methods section. What is striking, is how well all 4 estimations agree, especially when the incidence-per-day is high (more in the "Comparisons of $R_t$ Esimation Methods" section, below). 

The user is able to select a geographic region, and view a graph of how the $R_t$ changes over time for that region. Additionally, the user can view the cumulative number of cases and new cases per day for each region selected. The user can modify parameters associated with the $R_t$ calculation, such as specifying the size of the smoothing window (the number days used for a rolling mean) to smooth the number of new cases per day, the mean and standard deviation of the generation interval, and the number of days to consider when calculating $R_t$ for each day. 

For each geographic region select, the web-app summarizes the the last 3 weeks of $R_t$ into colored score cards. For each week, the score card displays the range of $R_t$ point estimates. The color is based on the average limits of the confidence interval (CI) of $R_t$ for a particular week: the mean upper limit is below 1 (green); the mean point estimate is below 1 but the mean upper limit is above 1 (yellow); the mean point estimate is above 1 but the mean lower limit is below 1 (orange); the mean lower limit is above 1 (red). 

## Comparisons of $R_t$ Esimation Methods

```{r echo = F, message=F, warning=F} 
library("tidyverse")
library("RcppRoll")
```

> **Box 1. Summary of comparison** <br>
> Using the data for US States, the 4 different methods result in very similar estimates of $R_t$. In particular when: <br>
> 1. Smoothing window is increased <br>
> 2. The number of days for assessing the time-period <br>
> 3. When the mean and standard deviation of generation interval are low <br>

```{r echo=F, message=F, warning=F}
df= read_csv("Fig2&3_data.csv")

data_RMSE_stats <- data_RMSE %>%
  filter(!is.infinite(RMSE)) %>%
  group_by(Comparison) %>%
  summarise(Mean = mean(RMSE, na.rm=T), 
            Median = median(RMSE, na.rm=T),
            `95_quantile` = quantile(RMSE, 0.95, na.rm=T))
```

When using the web-app, visual inspection shows that there is a great deal of agreement between the different methods of estimating $R_t$ (Figure 1 - add in a screenshot). We can quantify the difference in $R_t$ estimations using the absolute error. Here, we compared the $R_t$ values for each day of pandemic for 10 US States using the default parameter ^[smoothing window of 7 days, generation interval with mean of 4 day and standard deviation of 3 days, and time window for $R_t$ calculation of 7 days.], and we find very similar estimates of $R_t$ between the 4 methods - the 95th percentile of absolute error remains less that 0.7 (Figure 2). 

In particular, we find that as the incidents increase with the time period used to calculate $R_t$ (default 7 days), the difference in $R_t$ estimations between the different methods decreases (Figure 3). Similarly, the disagreement between the methods decreases if the size of the smoothing window is increased (Figure 4) or if the number of days used to calculate $R_t$ is increased (Figure 5). 

An important parameters when caculating $R_t$ is the distribution of the generation interval. We find as the mean or standard deviation in the generation interval increased, there is also increased disagreement in $R_t$ estimations between the four methods (Figure 6).

```{r echo=F, message = F, warning = F, fig.cap="Figure 2. Distribution of absolute error between all pairs of methods for estimating R", fig.width=12}

library("ggpubr")
ggplot(data_RMSE) + 
  geom_histogram(aes(RMSE), fill="white", color="black") +
  geom_vline(data = data_RMSE_stats, aes(xintercept = Median), linetype = 2) +
  geom_text(data = data_RMSE_stats, aes(x = Median*0.5, y = 150, 
                                        label = paste("Median:", round(Median,2)),
                                        hjust = 1)) + 
  geom_text(data = data_RMSE_stats, aes(x = `95_quantile`*1.5, y = 150, 
                                        label = paste("95th Percentile:", round(`95_quantile`,2)),
                                        hjust = 0)) + 
  geom_vline(data = data_RMSE_stats, aes(xintercept = `95_quantile`), linetype = 2) +
  labs(x = "Absolute Error") + 
  scale_x_log10(labels = scales::comma, limits=c(0.0001, 100)) +
  facet_wrap(~Comparison) + 
  theme_pubr()
```

```{r echo=F, message = F, warning = F, fig.cap = "Figure 3. The absolute error decreases as incidence increases.", fig.width=12}
library("ggpubr")

df <- df %>%
  group_by(region) %>%
  arrange(date) %>%
  mutate(weekly_sum = roll_sum(new_cases, n=7, align="right", fill = c(NA, NA, NA)))

data_RMSE <- df %>% 
  pivot_longer(cols = starts_with("RMSE"), names_to = "Comparison", values_to = "RMSE") %>%
  mutate(Comparison = factor(Comparison, levels = c("RMSE.KN_Cori", "RMSE.KN_WT", "RMSE.KN_WL", 
                                                    "RMSE.Cori_WT", "RMSE.Cori_WL", "RMSE.WT_WL")))

data_RMSE$Comparison <- data_RMSE$Comparison %>% fct_recode(`Cori vs Wallinga & Lipsitch` = "RMSE.Cori_WL",
                                                            `Cori vs Wallinga & Teunis` = "RMSE.Cori_WT",
                                                            `Simple Ratio vs Cori` = "RMSE.KN_Cori",
                                                            `Simple Ratio vs Wallinga & Lipsitch` = "RMSE.KN_WL",
                                                            `Simple Ratio vs Wallinga & Tuenis` = "RMSE.KN_WT",
                                                            `Walinga & Teunis vs Wallinga & Lipsitch` = "RMSE.WT_WL")

data_RMSE %>% filter(RMSE < 100) %>%
ggplot() +  
  geom_point(aes(new_cases_smoothed, RMSE), alpha=0.2) +
  facet_wrap(~Comparison, scales="free") +
  labs(x = "New cases per day", y = "Absolute Error") + 
  scale_y_log10(labels = scales::comma) + scale_x_log10(labels = scales::comma)

```

```{r echo=F, message = F, warning = F, fig.cap = "Figure 4. The absolute error decreases as the number days used to smooth new cases per day increases.", fig.width=12}

df = read_csv("Fig4_data.csv")
ggplot(df) +
  geom_line(aes(smoothing_window, error, color=Comparison)) +
  facet_wrap(~error_type, scales="free") + 
  theme_pubr()
```

```{r echo=F, message = F, warning = F, fig.cap = "Figure 5. The absolute error decreases as the number days used to calculate $R_t$ is increased", fig.width=12}

df = read_csv("Fig5_data.csv")
ggplot(df) +
  geom_line(aes(tau, error, color=Comparison)) +
  facet_wrap(~error_type, scales="free") + 
  theme_pubr()
```

```{r echo=F, message = F, warning = F, fig.cap = "Figure 6. The absolute error tends to increase as the mean and standard deviation of the generation interval are increased. (A) Median Absolute Error. (B) 95th Percentile in Absolute Error.", fig.width=12}

df = read_csv("Fig6_data.csv")

a = ggplot(df) +
  geom_raster(aes(x=mean, y=SD, fill=Median)) +
  facet_wrap(~Comparison) + 
  scale_x_continuous(breaks = c(2,4,6,8,10)) + 
  scale_y_continuous(breaks = c(2,4,6,8,10)) + 
  scale_fill_continuous(type = "viridis") +
  theme_pubr()


b = ggplot(df) +
  geom_raster(aes(x=mean, y=SD, fill=`95th_percentile`)) +
  facet_wrap(~Comparison) + 
  scale_x_continuous(breaks = c(2,4,6,8,10)) + 
  scale_y_continuous(breaks = c(2,4,6,8,10)) + 
  scale_fill_continuous(type = "viridis") +
  theme_pubr()

ggarrange(a,b, ncol=2, labels=c("A", "B"))
```