---
title: "Results"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
header-includes:
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The $R_t$ Estimator Web App

We developed a Shiny web-app that allows users to estimate $R_t$ for the SARS-CoV-2 pandemic in all countries, all states in the United States,, and all counties in the United States over time with 95% confidence intervals (or credible intervals) for the estimations. The user can estimate R using 4 different methods, all based on new cases per day and the generation interval (Table 1.) 

**Table 1. Methods for Estimating $R_t$**. 

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
Method                           Description                                                                               Equation                                            Reference                                    
-------------------------------- ----------------------------------------------------------------------------------------- --------------------------------------------------- ---------------------------------------------
**Ratio Model**                  $R_t$ is the sum of new cases ($I$) within a time window (size $\tau$) centered on day    $R_t = \frac{\sum_{i = t - \frac{\tau}{2} + D}^{t                                                
                                 $t + D$ divided by the sum of new cases on a time window centered on day $t$, where $D$   + \frac{\tau}{2} + D}I_i}{\sum_{i = t -                                                          
                                 is the mean generation interval.                                                          \frac{\tau}{2}}^{t + \frac{\tau}{2}}I_i}$                                                        

**Infectiousness Model**         $R_t$ is calculated as the number of new cases ($I$) within a time window (size $\tau$)   $R_t = \frac{\sum_{i = t                            Modified from Cori et al, 2013               
                                 centered on day $t$ divided by the sum of infectiousness of all individual infected       -\frac{\tau}{2}}^{t+\frac{\tau}{2}}I_i}{\sum_{i =                                                
                                 prior to the time window, where infectiousness is defined by the probability              0}^{t -\frac{\tau}{2}}I_i*w(i)}$                                                                 
                                 distribution of the generation interval                                                                                                                                                    

**Transmission Network Model**   For this method, all transmission networks compatible with the observed number of new     $p_{i,j} = \frac{w(t_i - t_j)}{\sum_{i \neq         Wallinga & Teunis, 2004; Obadia et al, 2012  
                                 cases per day are considered and a pre-defined distribution of generational intervals     k}{w(t_i - t_k)}}$ <br> $R_j = \sum_{i} p_{i,j}$                                                 
                                 ($w(t)$). The likelihood that a new case $i$ was infected by existing case $j$ is the     <br> $R_t = \bar{R_j}$                                                                           
                                 probability of a generation interval equal to the time difference between $i$ and $j$                                                                                                      
                                 ($w(t_i - t_j)$). The relative likelihood of $i$ being infected by $j$ ($p_{ij}$) is                                                                                                       
                                 $w(t_i - t_j)$ normalized to the likelihood of case $i$ having been infected by all                                                                                                        
                                 other cases $k$. The likely number of cases produced by $j$ ($R_j$), is the sum of                                                                                                         
                                 $p_{i,j}$ for all new cases $i$. Thus, $R_t$ is is the average of all $R_j$.                                                                                                               

**Exponential Growth Method**    $R_t$ is derived from rate exponential growth ($r$) observed within a time window (size   $R_t = \frac{1}{\int_{t = 0}^{\infty}               Modified from Wallinga & Lipsitch (2007)     
                                 $\tau$) centered on day $t$                                                               e^{-rt}w(t)dt}$                                                                                  
----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

The user is able to select one or more geographic regions, and view a graph of how the $R_t$ changes over time for each of those regions. The user can modify parameters associated with the $R_t$ calculation, such as specifying the size of the smoothing window (the number days used for a rolling mean) to smooth the number of new cases per day, the mean and standard deviation of the generation interval, and the number of days to consider when calculating $R_t$ for each day. Additionally, the user can view the cumulative number of cases and new cases per day for each region selected. 

For each geographic region select, the web-app summarizes the the last 3 weeks of $R_t$ into colored score cards. For each week, the score card displays the range of $R_t$ point estimates. The color is based on the average limits of the confidence interval (CI) of $R_t$ for a particular week: the mean upper limit is below 1 (green); the mean point estimate is below 1 but the mean upper limit is above 1 (yellow); the mean point estimate is above 1 but the mean lower limit is below 1 (orange); the mean lower limit is above 1 (red). 

## Comparisons of $R_t$ Esimation Methods

```{r echo = F, message=F, warning=F} 
library("tidyverse")
library("RcppRoll")
```

> **Box 1. Summary of comparison** <br>
> Using the data for US States, the 4 different methods result in very similar estimates of $R_t$. In particular when: <br>
> 1. Smoothing window is increased <br>
> 2. The number of days for assessing the time-period <br>
> 3. When the mean and standard deviation of generation interval are low <br>

When using the web-app, visual inspection shows that there is a great deal of agreement between the different methods of estimating $R_t$ (Figure 1 - add in a screenshot). We can quantify the difference in $R_t$ estimations using the absolute error. Here, we compared the $R_t$ values for each day of pandemic for 10 US States using the default parameter ^[smoothing window of 7 days, generation interval with mean of 4 day and standard deviation of 3 days, and time window for $R_t$ calculation of 7 days.], and we find very similar estimates of $R_t$ between the 4 methods - the 95th percentile of absolute error remains less that 0.7 (Figure 2). 

In particular, we find that as the incidents increase with the time period used to calculate $R_t$ (default 7 days), the difference in $R_t$ estimations between the different methods decreases (Figure 3). Similarly, the disagreement between the methods decreases if the size of the smoothing window is increased (Figure 4) or if the number of days used to calculate $R_t$ is increased (Figure 5). 

An important parameters when caculating $R_t$ is the distribution of the generation interval. We find as the mean or standard deviation in the generation interval increased, there is also increased disagreement in $R_t$ estimations between the four methods (Figure 6).

```{r echo=F, message = F, warning = F, fig.cap="Figure 2. Distribution of absolute error between all pairs of methods for estimating R", fig.width=12}

df= read_csv("DATA - baseline_parameters.csv")

data_RMSE_stats <- df %>%
  filter(!is.infinite(RMSE)) %>%
  group_by(Comparison) %>%
  summarise(Mean = mean(RMSE, na.rm=T), 
            Median = median(RMSE, na.rm=T),
            `95_quantile` = quantile(RMSE, 0.95, na.rm=T))

library("ggpubr")
ggplot(df) + 
  geom_histogram(aes(RMSE), fill="white", color="black") +
  geom_vline(data = data_RMSE_stats, aes(xintercept = Median), linetype = 2) +
  geom_text(data = data_RMSE_stats, aes(x = Median*0.5, y = 150, 
                                        label = paste("Median:", round(Median,2)),
                                        hjust = 1)) + 
  geom_text(data = data_RMSE_stats, aes(x = `95_quantile`*1.5, y = 150, 
                                        label = paste("95th Percentile:", round(`95_quantile`,2)),
                                        hjust = 0)) + 
  geom_vline(data = data_RMSE_stats, aes(xintercept = `95_quantile`), linetype = 2) +
  labs(x = "Absolute Error") + 
  scale_x_log10(labels = scales::comma, limits=c(0.0001, 100)) +
  facet_wrap(~Comparison) + 
  theme_pubr()
```

```{r echo=F, message = F, warning = F, fig.cap = "Figure 3. The absolute error decreases as incidence increases.", fig.width=12}
library("ggpubr")

df <- df %>%
  group_by(region) %>%
  arrange(date) %>%
  mutate(weekly_sum = roll_sum(new_cases, n=7, align="right", fill = c(NA, NA, NA)))

df %>% filter(RMSE < 100) %>%
ggplot() +  
  geom_point(aes(new_cases_smoothed, RMSE), alpha=0.2) +
  facet_wrap(~Comparison, scales="free") +
  labs(x = "New cases per day", y = "Absolute Error") + 
  scale_y_log10(labels = scales::comma) + scale_x_log10(labels = scales::comma)

```

```{r echo=F, message = F, warning = F, fig.cap = "Figure 4. The absolute error decreases as the number days used to smooth new cases per day increases.", fig.width=12}

df = read_csv("DATA - vary_smoothing_parameter.csv")

df <- df %>% group_by(smoothing_window, Comparison) %>%
  summarise(Median = median(RMSE, na.rm=T),
            `95th_percentile` = quantile(RMSE, 0.95, na.rm=T))

df$smoothing_window = as.numeric(df$smoothing_window)

df = df %>% pivot_longer(cols = c("Median", "95th_percentile"), names_to = "error_type", values_to = "error")
df$error_type =  df$error_type %>% fct_recode(`Median Error` = "Median",
                                              `95th Percentile of Error` = "95th_percentile")

ggplot(df) +
  geom_line(aes(smoothing_window, error, color=Comparison)) +
  facet_wrap(~error_type, scales="free") + 
  theme_pubr()
```

```{r echo=F, message = F, warning = F, fig.cap = "Figure 5. The absolute error decreases as the number days used to calculate $R_t$ is increased", fig.width=12}

df = read_csv("DATA - vary_tau_parameter.csv")

df <- df %>% group_by(tau, Comparison) %>% summarise(Median = median(RMSE, na.rm=T),
                                                     `95th_percentile` = quantile(RMSE, 0.95, na.rm=T))

df = df %>% pivot_longer(cols = c("Median", "95th_percentile"), names_to = "error_type", values_to = "error")
df$error_type =  df$error_type %>% fct_recode(`Median Error` = "Median",
                                              `95th Percentile of Error` = "95th_percentile")

ggplot(df) +
  geom_line(aes(tau, error, color=Comparison)) +
  facet_wrap(~error_type, scales="free") + 
  theme_pubr()
```

```{r echo=F, message = F, warning = F, fig.cap = "Figure 6. The absolute error tends to increase as the mean and standard deviation of the generation interval are increased. (A) Median Absolute Error. (B) 95th Percentile in Absolute Error.", fig.width=12}

df = read_csv("DATA - vary_GT_parameter.csv")

df <- df %>% group_by(GT_mean, GT_SD, Comparison) %>% summarise(Median = median(RMSE, na.rm=T),
                                                    `95th_percentile` = quantile(RMSE, 0.95, na.rm=T))

a = ggplot(df) +
  geom_raster(aes(x=GT_mean, y=GT_SD, fill=Median)) +
  labs(x = "Mean", y="SD") +
  facet_wrap(~Comparison) + 
  scale_x_continuous(breaks = c(2,4,6,8,10)) + 
  scale_y_continuous(breaks = c(2,4,6,8,10)) + 
  scale_fill_continuous(type = "viridis") +
  theme_pubr()

b = ggplot(df) +
  geom_raster(aes(x=GT_mean, y=GT_SD, fill=`95th_percentile`)) +
  labs(x = "Mean", y="SD") +
  facet_wrap(~Comparison) + 
  scale_x_continuous(breaks = c(2,4,6,8,10)) + 
  scale_y_continuous(breaks = c(2,4,6,8,10)) + 
  scale_fill_continuous(type = "viridis") +
  theme_pubr()

ggarrange(a,b, ncol=2, labels=c("A", "B"))
```