---
title: "Code for R Estimator"
output: html_document
---

```{r setup, include=FALSE, echo=F, message=F, warning=F}
knitr::opts_chunk$set(echo = TRUE)
```

## Download Data

```{r download_data, message=F, warning=F}
library("tidyverse")

#download county level data from NYT 
US_countydata <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv", col_types = "Dcccnn")

#download country level data from covid.ourworldindata.org
countrydata <- read_csv("https://covid.ourworldindata.org/data/ecdc/full_data.csv", col_types = "Dcnnnn")

#download US State level data from covidtracking.com
column_specs = cols(
    date = col_date(format = "%Y%m%d"),
    state = col_character(),
    positive = col_double(),
    negative = col_double(),
    pending = col_double(),
    hospitalizedCurrently = col_double(),
    hospitalizedCumulative = col_double(),
    inIcuCurrently = col_double(),
    inIcuCumulative = col_double(),
    onVentilatorCurrently = col_double(),
    onVentilatorCumulative = col_double(),
    recovered = col_double(),
    dataQualityGrade = col_character(),
    lastUpdateEt = col_datetime(format = "%m/%d/%Y %H:%M"),
    hash = col_character(),
    dateChecked = col_skip(),
    death = col_double(),
    hospitalized = col_double(),
    total = col_double(),
    totalTestResults = col_double(),
    posNeg = col_double(),
    fips = col_character(),
    deathIncrease = col_double(),
    hospitalizedIncrease = col_double(),
    negativeIncrease = col_double(),
    positiveIncrease = col_double(),
    totalTestResultsIncrease = col_double()
)

US_statedata <- read_csv("https://covidtracking.com/api/v1/states/daily.csv", col_types = column_specs)

```


## Standardize Data

```{r standardize_data, warning=F, message=F}
library(tigris)

# Create a unique label for each region in each dataset

US_countydata <- US_countydata %>%
    unite(col = region, county, state, sep=", ") %>%
    dplyr::select(date, region, regionID = fips, total_cases = cases) %>%
    mutate(region_type = "county", regionID_type = "fips")

US_statedata <- US_statedata %>%
    dplyr::select(date, region = state, total_cases = positive, new_cases = positiveIncrease, regionID = fips) %>%
    mutate(region_type = "state", regionID_type = "fips")
    countrydata <- countrydata %>%
    dplyr::select(date, region = location, new_cases, total_cases) %>%
    mutate(region_type = "nation", regionID = NA, regionID_type = NA)
    
#Replace abberviated names with full names in US State data
fips_reference = tigris::fips_codes %>% dplyr::select(1:3) %>% distinct()
    US_statedata = left_join(US_statedata, fips_reference, by=c("regionID" = "state_code")) %>%
    mutate(region = state_name) %>%
    mutate(region = paste(region, "United States", sep=", ")) %>%
    dplyr::select(-c(state, state_name))
```

<div id="bind_data">
Bind rows together
```{r bind_data}
DATA <- bind_rows(list(US_countydata, US_statedata, countrydata))
```
</div>

## Clean Data

<div id="clean_1">
For each region, remove leading dates that are just a series of 0s or NAs 

```{r clean_step1, warning=F, message=F}
DATA2 <- DATA %>%
  filter(!is.na(total_cases)) %>%
  group_by(region, regionID, region_type, regionID_type) %>%
  summarise(reference_date = min(date))

DATA <- full_join(DATA, DATA2) %>%
  filter(date >= reference_date) %>%
  dplyr::select(-reference_date)
```

</div>

<div id="clean_2">
Sometimes there are errors in the reporting of new cases and when the error is discovered, the cumulative number of cases decreases. To correct for this, for each region, remove dates that have cumulative number of cases is greater than a future date. 

```{r clean_step2, warning=F, message=F}

#Group data by region and arrange by date
DATA <- DATA %>%
  group_by(region, regionID, region_type, regionID_type) %>%
  arrange(date)

# Calculate new cases per day per region
DATA <- DATA %>% mutate(new_cases = c(0, diff(total_cases)))

# Create a column that will be used to mark dates that will be needed to interprolated (see next code block on interprolation)
DATES_TO_INTERPROLATE = filter(DATA, F) %>% ungroup()

DATA = DATA %>%
  # identify days with negative number of "new cases"
  mutate(negative_increase = new_cases < 0) %>%
  mutate(has_negative_newcases = any(negative_increase, na.rm=T))

while(any(DATA$new_cases < 0)) {
  DATA = split(DATA, DATA$has_negative_newcases)
  
  DATA[["TRUE"]] = DATA[["TRUE"]] %>%
    # identify when the latest date in which new_cases < 0
    mutate(reference_date = which.max(negative_increase)) %>%
    # identify days before reference_date that have more total_cases than reference_date
    mutate(rows_to_filter = (((date < date[reference_date]) & (total_cases > total_cases[reference_date]))))
  
  # add days you will filter out to DATES_TO_INTERPROLATE
  DATES_TO_INTERPROLATE = DATA[["TRUE"]] %>% filter(rows_to_filter == T) %>% bind_rows(DATES_TO_INTERPROLATE, .)
  
  DATA[["TRUE"]] = DATA[["TRUE"]] %>%
    # filter out days
    filter(rows_to_filter == F) %>%
    # recalculate new_cases
    mutate(new_cases = c(0, diff(total_cases)))
  
  DATA = bind_rows(DATA)
  
  DATA = DATA %>%
    # identify days with negative number of "new cases"
    mutate(negative_increase = new_cases < 0) %>%
    mutate(has_negative_newcases = any(negative_increase))
}

# Remove that are no longer necessary
DATA = DATA %>%
  dplyr::select(-c(negative_increase, has_negative_newcases, reference_date, rows_to_filter)) %>%
  ungroup()
```

</div>

<div id="clean_3">
For the dates that are removed in the above step, we will interporlate the cumulative number of cases as linearly increasing between the two dates that have known cumulative case numbers.  

```{r clean_step3, message=F, warning=F}  
# -------------------------------------------------------
# Interprolate total_cases for dates that were removed
# -------------------------------------------------------
library(zoo)

DATES_TO_INTERPROLATE = DATES_TO_INTERPROLATE %>%
  dplyr::select(region, region_type, regionID, regionID_type, date) %>%
  mutate(INTERPROLATED = T)
  DATA = DATA %>% mutate(INTERPROLATED = F)

DATA = bind_rows(DATA, DATES_TO_INTERPROLATE)
DATA = DATA %>% dplyr::select(region, region_type, regionID, regionID_type, date, total_cases, new_cases, INTERPROLATED)

DATA = DATA %>%
group_by(region, region_type, regionID, regionID_type) %>%
arrange(date) %>%
mutate(total_cases = zoo::na.approx(total_cases, na.rm=F)) %>%
mutate(new_cases = c(0, diff(total_cases)))

# -------------------------------------------------------
# Replace NA values in new_cases with 0
# -------------------------------------------------------

DATA$new_cases = replace(DATA$new_cases, which(is.na(DATA$new_cases)), 0)
```

</div>

```{r echo=F, message=F, warning=F}
write_csv(DATA, "case_data.csv")
```