---
title: "Comparing Simplified estimation of R to EpiEstim estimation of R"
author: "Arya Zandvakili"
date: "5/7/2020"
output:
  html_document: default
    fig_caption: yes
  pdf_document: default
    fig_caption: yes
urlcolor: blue
---

## Introduction 

EpiEstim is an R^[1] package  for estimating the effective reproduction number that was developed by [Cori et al, 2013](https://academic.oup.com/aje/article/178/9/1505/89262). In this document we compare the two approaches. 

--- 
1. "R" referring to the coding package, it is a coincidence that it has the same name as the reproduction number]

## Section 1: An Analytical Comparison

They use a Bayesian approach to estimate R. They estimate R as gamma-distributed:

$$ R_t \sim \Gamma(a + \sum_{s=t-\tau+1}^{t} I_s, \frac{1}{\frac{1}{b} + \sum_{s=t-\tau+1}^{t} I_sw_s})$$

$$ mean(R_t) = \frac{a + \sum_{s=t-\tau+1}^{t} I_s}{\frac{1}{b} + \sum_{s=t-\tau+1}^{t} I_sw_s}$$

where $I_s$ is the number of new cases at time s, $\tau$ is the window size (i.e. in our case, we chose 7 as default), and $w_s$ is a probability distribution of how infectious individuals are at time $t$ given that they were infected $t-s$ days ago (i.e. this is the probability distribution of the serial interval). Also, $a$, and $b$ are the parameters for the Gamma prior distributions (default is $a$ = 1, and $b$ = 5, or in other words the prior mean and standard deviation for $R_t$ are 5 and 5, respectively). 

In comparison, the simplified calculation for the reproduction number at time $t$ is: 

$$ R_t^s = \frac{\sum_{s=t-\tau+1}^{t} I_s}{\sum_{s=t-\tau+1-D}^{t-D} I_s} $$

where $D$ is the estimated serial interval (or generation interval). 

Therefore, it seems to me that $R_t^S$ approximates $R_t$ so long as $\sum I_s >> a$ and the majority of the mass of the $w_s$ is at $D$. This is consistent with what we see in the web-app: $R_t^S$ and $R_t$ converge as the number of new cases per day increases, and in the EpiEstim controls we set the mean serial interval to equal D and we keep the standard deviation of the serial interval low. 

## Section 2: A Statistical Comparison 

Here we will compare the two approaches to estimating R based on the available data for the US states: 

#### 2.1 Load the data and calculate new cases per day. 

```{r message=F, warning=F}
library('tidyverse')

# *****************
# Download coronavirus case data for US States from NYT
# *****************

data <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv", col_types = "Dccnn")

# *****************
# Calculate new cases per day for each state
# *****************
# New_Cases_(i) = Total_Cases_(i) - Total_Cases_(i-1)
#     Where (i) is the day
# Total_Cases_(i) = Active_Cases_(i) - Deaths_(i)
#     The issue with this that 

data <- data %>%
    # remove unncessary column
    select(-fips) %>% 
    # calculate total cases
    mutate(total_cases = cases + deaths) %>%
    # group data by state and arrange by date
    group_by(state) %>% arrange(date) %>%
    # calculate new cases per day
    mutate(new_cases = total_cases - c(0, total_cases[-length(total_cases)])) %>%
    # ungroup
    ungroup()

# *****************
# Output
# *****************

DT::datatable(data)
```

#### 2.2 Estimate R via both methods
```{r message=F, warning=F}

var_tau = 7 # Window size
var_D = 4  # Mean serial interval 
var_D_sd = 1 # SD in serial interval

# *****************
# Estimate R using simplified method
# *****************

data <- data %>%
  group_by(state) %>%
  arrange(date) %>%
  mutate(new_cases.weekly_sum = roll_sum(new_cases, n=var_tau, align="right", fill = c(NA, NA, NA))) %>%
  mutate(denominator = head(c(rep(NA, var_D), new_cases.weekly_sum), var_D*-1)) %>%
  ungroup() %>%
  mutate(denominator = replace(denominator, which(denominator == 0), NA)) %>%
  mutate(Rs = new_cases.weekly_sum/denominator)

# *****************
# Estimate R using EpiEstim method
# *****************

library("EpiEstim")
estimate_R.safely = purrr::safely(estimate_R)


data = data %>%
  arrange(date) %>%
  group_by(state) %>%
  nest(CD = -state) %>%
  mutate(EpiEstim = map(CD, function(CD_)
      estimate_R.safely(CD_$new_cases, 
             method = "parametric_si", 
             config = make_config(list(mean_si = var_D, 
                                       std_si = var_D_sd,
                                       t_start = seq(2, nrow(CD_) - var_tau + 1),
                                       t_end = seq(var_tau+1, nrow(CD_))))))) %>%
  mutate(EpiEstim = map(EpiEstim, function(i) i$result$R)) %>%
  filter(!is.null(EpiEstim[[1]])) %>%
  mutate(CD = map(CD, function(CD_) tail(CD_, var_tau*-1))) %>%
  mutate(results = map2(CD, EpiEstim, cbind)) %>%
  select(state, results) %>%
  unnest(results)

# *****************
# Calculate Difference in R
# *****************

data <- data %>%
  mutate(Std_Error = (`Mean(R)` - Rs)/`Mean(R)`) %>%
  mutate(RMSE = sqrt((`Mean(R)` - Rs)^2))

# *****************
# Show data
# *****************

DT::datatable(data)

```

#### 2.3 Results

```{r continuous_error message=F, warning=F, echo=F, fig.cap="\label{fig:figs} Error in R estimates. (A) Comparing absolute values of R estimations by Simplified method and EpiEstim method. Dotted line is y=x. (B) Histogram of difference between EpiEstim R and Simplified R estimations. (C) Histogram of Root-Mean-Square-Error as a percent of the EpiEstim R calculation. (D) Root-Mean-Square-Error as a percent of the EpiEstim R calculation over each day of the epidemic. Each line represents one US State."}

library("ggpubr")

p1.a = data %>%
  ggplot(aes(`Mean(R)`, Rs)) + 
  geom_point(alpha=0.2) + 
  geom_abline(slope = 1, intercept = 0, linetype = 2) + 
  labs(x="EpiEstim R", y="Simplified R") +
  theme_pubr() +
  scale_x_log10() + scale_y_log10()

p1.b = data %>%
  mutate(Std_Error = map_dbl(Std_Error, function(i) ifelse(i > 2, 2, i))) %>%
  mutate(Std_Error = map_dbl(Std_Error, function(i) ifelse(i < -2, -2, i))) %>%
  ggplot() + 
  geom_histogram(aes(Std_Error), fill="white", color="black") + 
  geom_vline(xintercept = 0, linetype=2) + 
  labs(x = "R - Rs") + 
  theme_pubr() + 
  annotate(geom = "text", x=-1, y=1500, label = paste("Median Error =", round(median(data$Std_Error, na.rm=T),3))) + 
  annotate(geom = "text", x=-1, y=1000, label = paste("Mean Error =", round(mean(data$Std_Error, na.rm=T),3)))

p2.a = data %>%
  ggplot() + 
  geom_histogram(aes(100*RMSE/`Mean(R)`), fill="white", color="black") + 
  scale_x_log10(labels = scales::comma) +
  labs(x = "Root-Mean-Square-Error(%)") +
  theme_pubr() +
  annotate(geom = "text", x=0.1, y=700, label = paste("Median RMSE =", round(median(data$RMSE, na.rm=T),3))) + 
  annotate(geom = "text", x=0.1, y=600, label = paste("Mean RMSE =", round(mean(data$RMSE, na.rm=T),3)))
  
p2.b = data %>%
  group_by(t_end) + 
  summarise(RMSE.mean = mean(100*RMSE/`Mean(R)`)) + 
  ggplot() + 
  geom_line(data=data, aes(x = t_end, y = 100*RMSE/`Mean(R)`, group=state), alpha=0.25) +
  geom_line(aes(x = t_end, y = 100*RMSE/`Mean(R)`, group=state), alpha=0.25) +
  labs(x = "days since first case", y = "Root-mean-square-error (%)") + 
  scale_y_log10(labels = scales::comma) + 
  theme_pubr() + 
  theme(legend.position = "none")


ggarrange(p1.a, p1.b, p2.a, p2.b, 
          labels = c("A", "B", "C", "D"),
          ncol = 2, nrow = 2)
```

In the figure above we can see that the simplified R estimation levels off to be about 1% 
